---
title: 'Lecture 6: Visualization'
output:
  html_document:
    df_print: paged
---


```{r echo=FALSE,messages=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

### replotting Tufte's o-ring data

```{r}
oringDf=read.table('orings.csv',sep=',',header=TRUE)

ggplot(oringDf,aes(x=Temperature,y=DamageIndex)) +
  geom_point() + xlim(25,85) + ylab('Damage Index') +
  xlab('Temperature at time of launch') +
  geom_smooth(se=FALSE,span=1) + ylim(0,12) +
  geom_vline(xintercept=27.5,size=7,alpha=0.3,color='red')
```

### showing the data

Examples of a linear fit and the many ways it could be generated

```{r}
set.seed(1234567)
npts=12
df=data.frame(x=seq(1,npts))
df$yClean=df$x + rnorm(npts,sd=0.1)
pointSize=2
t=theme(axis.text=element_text(size=12),axis.title=element_text(size=16))
ggplot(df,aes(x,yClean)) + 
  geom_smooth(method='lm',se=FALSE) + ylim(-5,20) +
  ylab('Dental health') + xlab('Time spent flossing') + t
  

ggplot(df,aes(x,yClean)) + 
  geom_point(size=pointSize) +
  geom_smooth(method='lm',se=FALSE) + ylim(-5,20) +
  ylab('Dental health') + xlab('Time spent flossing') + t

df$yDirty=df$x+ rnorm(npts,sd=10)

ggplot(df,aes(x,yDirty)) + 
  geom_point(size=pointSize) +
  geom_smooth(method='lm',se=FALSE)+ ylim(-5,20) +
  ylab('Dental health') + xlab('Time spent flossing') + t

df$yOutlier=rnorm(npts)
df$yOutlier[npts]=200


ggplot(df,aes(x,yOutlier)) + 
  geom_point(size=pointSize) +
  geom_smooth(method='lm',se=FALSE) +
  ylab('Dental health') + xlab('Time spent flossing') + t

```

### Examples of good plotting

Show box plots and violin plots

```{r}
library(NHANES)
NHANES$isChild <- NHANES$Age<18

NHANES_adult=subset(NHANES,subset=!isChild & Height!='NA')

# plot height by Gender
dfmean=NHANES_adult %>% group_by(Gender) %>% summarise(Height=mean(Height))

ggplot(dfmean,aes(x=Gender,y=Height)) + 
  geom_bar(stat="identity",color='gray') + 
  coord_cartesian(ylim=c(0,210)) +
  ggtitle('Bar graph') + 
  theme(aspect.ratio=1)  

ggplot(dfmean,aes(x=Gender,y=Height)) + 
  geom_bar(stat="identity",color='gray') + 
  coord_cartesian(ylim=c(0,210)) +
  ggtitle('Bar graph') + 
  theme(aspect.ratio=1)  +
  geom_jitter(data=NHANES_adult,aes(x=Gender,y=Height),width=0.1,alpha=0.1)



ggplot(NHANES_adult,aes(x=Gender,y=Height)) + 
  geom_violin() + 
  ggtitle('Violin plot') + theme(aspect.ratio=1)

ggplot(NHANES_adult,aes(x=Gender,y=Height)) + 
  geom_boxplot() +  
  ggtitle('Box plot') + theme(aspect.ratio=1)

ggplot(subset(NHANES_adult,Gender=='female'),aes(Height)) +
  geom_histogram(binwidth = 1)

```

### Pie chart example

Pie chart was generated using Excel.  Compare to a bar chart

```{r}
religionData=read.table('religion.csv',sep=',')
names(religionData)=c('Religion','Percentage')
religionData = arrange(religionData,desc(Percentage))
religionData$Religion=factor(religionData$Religion,levels=religionData$Religion)
ggplot(religionData,aes(Religion,Percentage,label=Percentage)) +
  geom_bar(stat='identity') + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

### Effects of axis scaling

use data from https://www.ucrdatatool.gov/Search/Crime/State/RunCrimeStatebyState.cfm

```{r}
crimeData=read.table('CrimeStatebyState.csv',sep=',',header=TRUE)
crimeData = subset(crimeData,Year > 1989)
crimeData$ViolentCrimePerCapita=crimeData$Violent.crime.total/crimeData$Population
ggplot(crimeData,aes(Year,ViolentCrimePerCapita)) +
  geom_line() + ylim(-0.05,0.05)

ggplot(crimeData,aes(Year,ViolentCrimePerCapita)) +
  geom_line() 

ggplot(crimeData,aes(Year,ViolentCrimePerCapita)) +
  geom_line()  + ylim(0,0.01)

```

### Data-ink ratio

Examples of good or bad data/ink ratios

```{r}
ggplot(df,aes(x,yDirty)) + 
  geom_point(size=2) +
  geom_smooth(method='lm',se=FALSE)+ ylim(-5,20) +
  ylab('Dental health') + xlab('Time spent flossing')

ggplot(df,aes(x,yDirty)) + 
  geom_point(size=0.5) +
  geom_smooth(method='lm',se=FALSE)+ ylim(-5,20) +
  ylab('Dental health') + xlab('Time spent flossing') +
  theme(panel.grid.major = element_line(colour = "black",size=1)) +
  theme(panel.grid.minor = element_line(colour = "black",size=1)) 


```


### correction for inflation
Historical consumer price index data from https://inflationdata.com/Inflation/Consumer_Price_Index/HistoricalCPI.aspx?reloaded=true#Table
```{r}
cpiData=read.table('cpi_data.txt')
cpiDataTidy <- gather(cpiData, key, value, -V1)
head(cpiDataTidy)
meanCPIData=cpiDataTidy %>% group_by(V1) %>% summarize(meanCPI=mean(value))
ggplot(meanCPIData,aes(V1,meanCPI)) +
  geom_line()
```

Let's look at average cost of a gallon of gas, from http://www.thepeoplehistory.com/70yearsofpricechange.html

```{r}
gasPriceData=data.frame(year=c(1930,1940,1950,1960,1970,1980,1990,2009,2013),
                        gasPrice=c(.10,.11,.18,.25,.36,1.19,1.34,2.05,3.80))
# convert to 2013 dollars
cpiRef=meanCPIData[meanCPIData$V1==1950,]$meanCPI
gasPriceData$gasPrice1950Dollars=array(NA,dim(gasPriceData)[1])
for (i in 1:dim(gasPriceData)[1]){
  cpiYear=meanCPIData[meanCPIData$V1==gasPriceData$year[i],]$meanCPI
  gasPriceData$gasPrice1950Dollars[i]=gasPriceData$gasPrice[i]/(cpiYear/cpiRef)
}
ggplot(gasPriceData,aes(year,gasPrice)) +
  geom_line(aes(color='Nominal'),size=1) +
  geom_line(aes(year,gasPrice1950Dollars,color='Adjusted'),size=1) +
  ylab('Gasoline prices') + xlab('year') 
```

### zero on the Y axis

A demonstration of why one should always use zero as the base for bar graphs.

```{r}
ggplot(data.frame(y=c(100,95),x=c('condition 1','condition 2')),aes(x=x,y=y)) + 
  geom_col() 

ggplot(data.frame(y=c(100,95),x=c('condition 1','condition 2')),aes(x=x,y=y)) + 
  geom_col() + coord_cartesian(ylim=c(92.5,105))


```

