---
title: "Hypothesis testing - Live coding notebook"
output: html_notebook
---

```{r}
#Libraries
library(tidyverse)

#Parameters
set.seed(12345)
```

### Create a function that generates two groups of data with a given effect size and sample size

```{r}

# the sample size is per group (so the total will be 2 * sampleSize)
# effect size is in terms of standard deviation units

makeData <- function(effectSize,sampleSize){
  
  # first create a data frame containing 2 * sampleSize samples from a
  # standard normal distributin (mean = 0, sd = 1) in a variable called scores
  d <- tibble(scores = rnorm(2*sampleSize))
  
  # create a dummy variable called group that contains the values 0 for the first sampleSize 
  # elements, and 1 for the remaining elements
  # this is a bit tricky so we will give you the full code - the trick is to use the row numbers
  
  d <- d %>%
    rownames_to_column() %>%
    mutate(rownum = as.integer(rowname),
           group = as.integer(rownum > sampleSize)) %>%
    select(-rowname)
  
  # now add the effectSize to the rows in group 0
  d <- d %>%
    mutate(scores = scores + effectSize*group)
  
  # return a data frame that contains
  return(d)
}


```


### write our own function to perform a t-test using data generated with the previous function

```{r}

# take in a data frame like the one generated above
# assume that we testing the directional hypothesis that group==1 > group==0

myTTest <- function(d){
  # calculate the values that you need to compute the T statistic
  
  dSummary <- d %>%
    group_by(group) %>%
    summarize(groupMean = mean(scores),
              groupSD = var(scores),
              groupSize = n())
  
  # compute the T statistic
  t <- diff(dSummary$groupMean)/sqrt(dSummary$groupSD[1]/dSummary$groupSize[1] +
                                       dSummary$groupSD[2]/dSummary$groupSize[2])
  # compute the p-value
  p <- pt(t,df=nrow(d)-2,lower.tail = FALSE)
  return(c(t,p))
}

```

### Test it to make sure it gives us the same result as t.test()

```{r}

# make an example dataset
d = makeData(.5,36)

# compute t-test using our function
myResult <- myTTest(d)
myResult

# compute it using R's t.test
rResult <- t.test(scores ~ group,data=d,alternative='less',var.equal=TRUE)
rResult

```


### create a function that:
### uses the previous function to generate a dataset
### performs a t-test on that dataset to compare groups
### returns the p value for the test

```{r}

makeDataAndTest <- function(effectSize,sampleSize){
  # make the data using the specified effect size and sample size
  d <- makeData(effectSize,sampleSize)
  
  # perform the t-test
  ttestResult <- myTTest(d)
  
  # return the p-value
  return(ttestResult[2])
}
```

### create another function to run a simulation in which we perform the makeDataAndTest
function 1000 times and compute the proportion of significant results
- when the effect size is 0

```{r}

runSimulations <- function(nRuns = 1000, effectSize = 0, sampleSize = 36){
  simResults <- replicate(nRuns,makeDataAndTest(effectSize,sampleSize))

  pSignificant <- mean(simResults<.05)
  # need to return result as a data frame
  return(tibble(pSignificant=pSignificant))
}

runSimulations()
```

### run this function 1000 times while varying the effect size from 0 to 1 in steps of 0.1
- Holding sample size constant at 36

To do this we will use the dplyr do() verb


```{r}

simulateEffectSize <- tibble(effectSize = seq(0,1,0.1))

effectSizeResults <- simulateEffectSize %>% 
  group_by(effectSize) %>% 
  do(runSimulations(effectSize = .$effectSize))

```

### Plot the proportion of signficant results as a function of effect size

```{r}
ggplot(effectSizeResults,aes(effectSize,pSignificant)) +
  geom_line() + 
  xlab('Effect size (sd units)') +
  ylab('Proportion of significant results') +
  geom_hline(yintercept = .05,linetype='dashed') + 
  ylim(0,1)
```

### now run this function 1000 times while varying the sample size as specified
holding the effect size constant at 0.2 (a "small" effect)


```{r}

simulateSampleSize <- tibble(sampleSize = 2**seq(5,10))

sampleSizeResults <- simulateSampleSize %>% 
  group_by(sampleSize) %>% 
  do(runSimulations(sampleSize = .$sampleSize,effectSize=.2))

```

### Plot the proportion of signficant results as a function of sample size

```{r}
ggplot(sampleSizeResults,aes(sampleSize,pSignificant)) +
  geom_line() + 
  xlab('Sample size') +
  ylab('Proportion of significant results') +
  geom_hline(yintercept = .05,linetype='dashed') + 
  ylim(0,1) + 
  scale_x_continuous(breaks=sampleSizeResults$sampleSize)
```
