---
title: "More on data frames"
output:
  html_document:
    df_print: paged
---

One of the challenging aspects of coding in R is understanding data frames. This notebook provides a bit more detail and practice for working with them.

You can think of a data frame as a package that contains a number of diferent variables that are aligned with each other in rows.  For example, let's say that we wanted to create a variable to store height and free throw percentage for a set of basketball players.  First, let's create several vectors that contain the player's name, height, and free throw percentage.

Data obtained from https://www.teamrankings.com/nba/player-stat/free-throw-percentage


```{r}

name=c('Courtney Lee','Kevin Durant','Jamal Murray')
height=c(77,81,76)
freeThrowPercentage=c(94.4,88.9,92.4)

```

Now let's put these into a data frame.

```{r}
freeThrowDf = data.frame(name=name,
                         height=height,
                         freeThrowPercentage=freeThrowPercentage)
freeThrowDf
```

If we want to turn one of these variables back into a vector, we need to use the $ operation.

```{r}
freeThrowDf$freeThrowPercentage
```

Note that if we try to apply a function to the entire data frame, it won't work:

```{r}
mean(freeThrowDf)
```

But we can apply it to the individual variable:

```{r}
mean(freeThrowDf$freeThrowPercentage)
```

We can also access the data frame using indices that treat it like a two-dimensional vector (which we usually call a matrix).  We access this by specifying the row and column that we want, within square brackets.  The first number is the row, the second is the column.

```{r}
# first row, second column
freeThrowDf[1,2]

# second row, first column
freeThrowDf[2,1]
```

If you want all of the values, you can leave the index empty and it will give you all of them:

```{r}
# all rows, first column
freeThrowDf[,1]

# all columns, first row
freeThrowDf[1,]
```


