---
title: "More on data frames"
output:
  html_document:
    df_print: paged
---

One of the challenging aspects of coding in R is understanding data frames. This notebook provides a bit more detail and practice for working with them.

You can think of a data frame as a package that contains a number of diferent variables that are aligned with each other in rows.  For example, let's say that we wanted to create a variable to store height and free throw percentage for a set of basketball players.  First, let's create several vectors that contain the player's name, height, and free throw percentage.

Data obtained from https://www.teamrankings.com/nba/player-stat/free-throw-percentage


```{r}

name=c('Courtney Lee','Kevin Durant','Jamal Murray','Stephen Curry')
height=c(77,81,76,75)
freeThrowPercentage=c(94.4,88.9,92.4,91.3)

```

Now let's put these into a data frame.

```{r}
freeThrowDf = data.frame(name=name,
                         height=height,
                         freeThrowPercentage=freeThrowPercentage)
freeThrowDf
```


We can find out some information about the data frame using the dim function to see the dimensions, and the nrow and ncol functions to see the numbers of rows and columns respectively.

```{r}
dim(freeThrowDf)
nrow(freeThrowDf)
ncol(freeThrowDf)

```

If we want to turn one of these variables back into a vector, we need to use the $ operation.

```{r}
freeThrowDf$freeThrowPercentage
```

Note that if we try to apply a function that expects a vector to the entire data frame, it won't work:

```{r}
mean(freeThrowDf)
```

But we can apply it to the individual variable:

```{r}
mean(freeThrowDf$freeThrowPercentage)
```

We can also access the data frame using indices that treat it like a two-dimensional vector (which we usually call a matrix).  We access this by specifying the row and column that we want, within square brackets.  The first number is the row, the second is the column.

```{r}
# first row, second column
freeThrowDf[1,2]

# second row, first column
freeThrowDf[2,1]
```


If you want all of the values along one dimension, you can leave the index empty and it will give you all of them:

```{r}
# all rows, first column
freeThrowDf[,1]

# all columns, first row
freeThrowDf[1,]
```

