---
title: "Birthday example analysis"
output:
  html_document:
    df_print: paged
---

Here we examine a standard problem in probability theory known as the "birthday problem".  There are various versions, but the one that we examine is:  Given some number of students in the class, how many students do we expect to share a birthday with at least one other student?

```{r echo=FALSE,message=FALSE}
library(ggplot2)
```

We will use the data collected from the class using the Google Form.

```{r}
birthdayData=read.csv('Birthday form.csv')
nstudents=dim(birthdayData)[1]
birthdayData$Timestamp=NULL
names(birthdayData)=c('Birthday','Expected')

bdTable=table(birthdayData$Birthday)
print('Number of people who share a birthday with a classmate:')
nMatchesObserved=sum(bdTable[bdTable>1])
nMatchesObserved

print('Number of people you expected to share a birthday with a classmate:')

nMatchesPredicted=mean(birthdayData$Expected)*nstudents
nMatchesPredicted
```

Now let's see how many we would have expected. Let's simulate it first.

```{r}

nRuns=5000
nMatches=array(NA,nRuns)
for (i in 1:nRuns){
  # assign an integer from 1-365 to each individual
  # replacing after sampling
  birthdayDataSimulated=data.frame(birthday=sample.int(365,size=nstudents,replace=TRUE))
  bdTable=table(birthdayDataSimulated$birthday)
  nMatches[i]=sum(bdTable[bdTable>1])
}
print('expected number of matches:')
print(mean(nMatches))

ggplot(data.frame(nMatches=nMatches),aes(nMatches)) +
  geom_histogram(binwidth=1) +
  geom_vline(xintercept=nMatchesObserved, color='blue')
```

We can also compute it directly using probability theory.  Here I am adapting the explanation found at https://math.stackexchange.com/questions/35791/birthday-problem-expected-number-of-collisions:

The probability person B shares person A's birthday is $\frac{1}{N}$, where N is the number of equally possible birthdays,

so the probability B does not share person A's birthday is $1 - \frac{1}{N}$,

so the probability $n-1$ other people do not share A's birthday is $(1−\frac{1}{N})^{n−1}$,

so the expected number of people who do not have others sharing their birthday is $n*(1−\frac{1}{N})^{n−1}$,

so the expected number of people who share birthdays with somebody is $n*(1 - (1−\frac{1}{N})^{n−1})$.

```{r}
expected=(1 - (1-1/365)**(nstudents-1))*nstudents
print('expected number by probability theory:')
expected
```

