---
title: "Session 9: Probability"
output:
  html_document:
    df_print: paged
---

```{r echo=FALSE,message=FALSE}
library(ggplot2)
```

#### Sets in R

```{r}
setA=c(1,2,3)
setB=c(2,3,4,5)
union(setA,setB)
```

```{r}
intersect(setA,setB)

```


#### Probability distributions

# Steph Curry example

```{r}
pFreeThrows=dbinom(seq(0,4),4,0.91)
curryDf = data.frame(numSuccesses=seq(0,4),probability=pFreeThrows)
curryDf
sum(curryDf$probability[1:3])
```
Plot probability distribution

```{r}
ggplot(curryDf,aes(x=numSuccesses,y=probability)) +
  geom_bar(stat="identity") + scale_x_continuous(breaks=seq(0,4)) +
  xlab('Number of successes out of 10 free throws') + ylab('Probability') 
```


```{r}

ggplot(curryDf,aes(x=numSuccesses,y=probability)) +
  geom_bar(stat="identity") + scale_x_continuous(breaks=seq(0,4)) +
  xlab('Number of successes out of 10 free throws') + ylab('Probability') +
  geom_bar(data=curryDf[1:3,],aes(x=numSuccesses,y=probability),stat="identity",fill='green') 


```


Color in tail probability for 7+ heads - using cumulative

```{r}

curryDf$cumulative=cumsum(curryDf$probability)
ggplot(curryDf,aes(x=numSuccesses,y=cumulative)) +
  geom_bar(stat="identity") + scale_x_continuous(breaks=seq(0,4)) +
  xlab('Number of successes out of 10 free throws') + ylab('Cumulative Probability') +
  geom_bar(data=curryDf[1:3,],aes(x=numSuccesses,y=cumulative),stat="identity",fill='green') +
  geom_hline(yintercept=curryDf$probability[3],color='blue',linetype='dashed') 

```



#### Election results
Here we analyze the election results as they came in during the special election for US Senator in Alabama on Dec 12, 2017 (transcribed from https://www.ajc.com/news/national/alabama-senate-race-live-updates-roy-moore-doug-jones/KPRfkdaweoiXICW3FHjXqI/).  

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
electionReturns=read.table('alabama_election_returns.csv',sep=',',header=TRUE) %>% gather(candidate,pctVotes,-pctResp)
ggplot(electionReturns,aes(pctResp,pctVotes,color=candidate)) +
  geom_line(size=1) +
  scale_color_manual(values=c( "#9999CC","#CC6666")) +
  xlab('Percentage of precincts reporting') +
  ylab('Percentage of votes')

```

#### Law of large numbers

Let's run 100,000 trials of 10 coin flips, and look at how the average empirical probability changes as a fnction of sample size

```{r}
set.seed(12345)
nsamples=30000
sampDf=data.frame(x=seq(nsamples),samples=rbinom(nsamples,10,0.5)/10) %>% mutate(mean=cumsum(samples)/seq_along(samples))

ggplot(sampDf[10:nsamples,],aes(x=x,y=mean)) + 
  geom_hline(yintercept = 0.5,color='blue',linetype='dashed') +
  geom_line() + 
  xlab('Number of trials') + ylab('Estimated probability of heads')
```

