---
title: 'Session 15: Robust Statistics'
output:
  html_document:
    df_print: paged
---

In this session we examine the effects of outliers and how to fix them.  Let's use a sample of 100 individuals from the NHANES adult height data, but alter it so that there is an outlier; let's pretend that height for one of the individuals was accidentally reported in inches rather than centimeters.  Note that we sort the data so that the shortest individual is always the outlier; the results below do not hold exactly if we move an observation from the top to the bottom of the distribution.

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(NHANES)
library(assertthat)
sampSize=100

NHANES$isChild <- NHANES$Age<18
NHANES_adult=sample_n(subset(NHANES,select=c('Height'),subset=!isChild & Height!='NA'),sampSize) %>% arrange(Height)
NHANES_adult = NHANES_adult %>% mutate(HeightOutlier=Height)
NHANES_adult$HeightOutlier[1]=NHANES_adult$Height[1]/2.54 # convert original to inches

ggplot(NHANES_adult,aes(HeightOutlier)) +
  geom_histogram(bins=30)


```

Now let's see how the various measures of central tendency are affected by the outlier.

```{r}
summaryData=rbind(summarise(NHANES_adult,label='Original',mean=mean(Height),sd=sd(Height),median=median(Height)),
              summarise(NHANES_adult,label='Outlier',mean=mean(HeightOutlier),sd=sd(HeightOutlier),median=median(HeightOutlier)))
print(summaryData)
```

To see the effects across many possible samples, let's take 100 samples and compute a number of different measures of central tendency and dispersion with the outlier versus the original sample.

```{r}
library(boot)
library(MASS)

# we need this to use the boot package bootstrap function
samplemean <- function(x, d) {
  return(mean(x[d]))
}

getSample=function() {
  NHANES$isChild <- NHANES$Age<18
  NHANES_sample=sample_n(subset(NHANES,select=c('Height'),subset=!isChild & Height!='NA'),sampSize) %>% 
    arrange(Height)
  NHANES_sample$HeightOutlier=NHANES_sample$Height
  NHANES_sample$HeightOutlier[1]=NHANES_sample$HeightOutlier[1]/2.54
  return(NHANES_sample)
}

nSamples=100
sampleData=c()
for (i in 1:nSamples){
  s=getSample()
  bootResult=boot(s$Height, samplemean, R=1000)
  bootOutlierResult=boot(s$HeightOutlier, samplemean, R=1000)
  huberEst=huber(s$Height)
  huberEstOutlier=huber(s$HeightOutlier)
  results=cbind(summarise(s,mean=mean(Height),median=median(Height),
                        trimmedMean=mean(Height,trim=0.1),
                        bootMean=mean(bootResult$t),bootSd=sd(bootResult$t),
                        sd=sd(Height),mad=mad(Height),iqr=IQR(Height),huberMean=huberEst$mu,
                        huberMAD=huberEst$s),
                summarise(s,meanOutlier=mean(HeightOutlier),
                        medianOutlier=median(HeightOutlier),
                        trimmedMeanOutlier=mean(HeightOutlier,trim=0.1),
                        bootMeanOutlier=mean(bootOutlierResult$t),
                        bootSdOutlier=sd(bootOutlierResult$t),
                        sdOutlier=sd(HeightOutlier),
                        madOutlier=mad(HeightOutlier),
                        iqrOutlier=IQR(HeightOutlier),
                        huberMeanOutlier=huberEstOutlier$mu,
                        huberMADOutlier=huberEstOutlier$s))
  
  sampleData=rbind(sampleData,results)
}


```

Plot the results against one another

```{r}
ggplot(sampleData,aes(mean,meanOutlier,color='mean')) +
  geom_point() +
  geom_abline(slope=1,intercept=0) +
  xlim(165,172.5) +
  ylim(165,172.5) + 
  #geom_point(aes(median,medianOutlier,color='median'),alpha=0.5) +
  #geom_point(aes(trimmedMean,trimmedMeanOutlier,color='trimmedMean'),alpha=0.5) +
  xlab('Original data') + ylab('With outlier') +
  ggtitle(sprintf('mean difference = %f',mean(sampleData$mean-sampleData$meanOutlier)))

```

How about the trimmed mean?

```{r}
ggplot(sampleData,aes(trimmedMean,trimmedMeanOutlier,color='mean')) +
  geom_point() +
  geom_abline(slope=1,intercept=0) +
  xlim(165,172.5) +
  ylim(165,172.5) + 
  #geom_point(aes(median,medianOutlier,color='median'),alpha=0.5) +
  #geom_point(aes(trimmedMean,trimmedMeanOutlier,color='trimmedMean'),alpha=0.5) +
  xlab('Original data') + ylab('With outlier') +
  ggtitle(sprintf('mean difference = %f',mean(sampleData$trimmedMean-sampleData$trimmedMeanOutlier)))

NHANES_adult$trimmed=FALSE
NHANES_adult$trimmed[NHANES_adult$HeightOutlier<quantile(NHANES_adult$HeightOutlier,0.1)]=TRUE
NHANES_adult$trimmed[NHANES_adult$HeightOutlier>quantile(NHANES_adult$HeightOutlier,0.9)]=TRUE
ggplot(NHANES_adult,aes(HeightOutlier,fill=trimmed)) +
  geom_histogram(bins=30) 
  

```

How about the median?

```{r}
ggplot(sampleData,aes(median,medianOutlier,color='mean')) +
  geom_point() +
  geom_abline(slope=1,intercept=0) +
  xlim(165,172.5) +
  ylim(165,172.5) + 
  #geom_point(aes(median,medianOutlier,color='median'),alpha=0.5) +
  #geom_point(aes(trimmedMean,trimmedMeanOutlier,color='trimmedMean'),alpha=0.5) +
  xlab('Original data') + ylab('With outlier') +
  ggtitle(sprintf('mean difference = %f',mean(sampleData$median-sampleData$medianOutlier)))

```

Can bootstrapping help with outliers?

```{r}
ggplot(sampleData,aes(bootMean,bootMeanOutlier,color='mean')) +
  geom_point() +
  geom_abline(slope=1,intercept=0) +
  xlim(165,173) +
  ylim(165,173) + 
  xlab('Original data') + ylab('With outlier') + 
    ggtitle(sprintf('mean difference = %f',mean(sampleData$bootMean-sampleData$bootMeanOutlier)))


```

What about using an M-estimator?

```{r}
ggplot(sampleData,aes(mean,huberMeanOutlier,color='Huber M-estimate')) +
  geom_point() +
  geom_abline(slope=1,intercept=0) +
  xlim(165,175) +
  ylim(165,175) + 
  xlab('Original data') + ylab('With outlier')

```

We can also look at the effect of outliers on estimates of dispersion.

```{r}
ggplot(sampleData,aes(sd,sdOutlier,color='std deviation')) +
  geom_point() + xlim(10,18) + ylim(10,18) +
  geom_abline(slope=1,intercept=0) +
  geom_point(aes(iqr,iqrOutlier,color='IQR'),alpha=0.5) +
  xlab('Original data') + ylab('With outlier')

```

What about bootstrapping?


```{r}
sampleData['se']=sampleData$sd/sqrt(nSamples)
sampleData['seOutlier']=sampleData$sdOutlier/sqrt(nSamples)
ggplot(sampleData,aes(bootSd,bootSdOutlier,color='bootstrapped std error')) +
  geom_point() +
  geom_abline(slope=1,intercept=0) +
  geom_point(aes(se,seOutlier,color='std error'),alpha=0.5) +
  xlab('Original data') + ylab('With outlier') +
  xlim(0,4) + ylim(0,4)

```

And the M-estimator?
```{r}
ggplot(sampleData,aes(huberMAD,huberMADOutlier,color='Huber M-estimator MAD')) +
  geom_point() +
  geom_abline(slope=1,intercept=0) +
  xlab('Original data') + ylab('With outlier')

```

### Assertion

```{r}
maxHumanHeightCm=272
minHumanHeightCm=60
#assert_that(all(NHANES_adult$HeightOutlier>minHumanHeightCm))
#assert_that(all(NHANES_adult$HeightOutlier<maxHumanHeightCm))
```

# review: normal vs. long-tailed distributions

```{r}

fbdata=read.table('https://rawgit.com/psych10/psych10/master/notebooks/Session04-SummarizingData/facebook_combined.txt')
print(sprintf('found %d unique IDs',length(unique(fbdata[,1]))))
# use the table function to compute how many times each individual ID shows up in the dataset
friends_table=table(fbdata[,1])
nfriends=as.data.frame(friends_table)
names(nfriends)=c('ID','Nfriends')

ggplot(nfriends, aes(Nfriends)) +
  geom_histogram(aes(y=..density..),fill = "red",binwidth=2)  +
  xlab('Number of friends') 
```

Plot power function

```{r}
nfriends = nfriends %>% mutate(logNfriends = log10(Nfriends))

h=hist(nfriends$Nfriends,breaks=seq(min(nfriends$Nfriends)-0.5,max(nfriends$Nfriends)+0.5))
histDf=data.frame(nfriends=h$mids,
                  logNfriends=log10(h$mids),
                  number=h$counts,
                  logNumber=log10(h$counts),
                  sqrtNumber=sqrt(h$counts))

histDf= histDf %>% subset(number>0)
logfit=lm(logNumber ~ logNfriends,data=histDf)
print(logfit)
histDf$logPredicted=10**predict(logfit)*sum(h$counts)

ggplot(histDf, aes(x=nfriends,y=number)) +
  geom_point()  +
  xlab('Number of friends')  + 
  geom_line(aes(x=nfriends,y=logPredicted),color='blue') +
  ylab('Frequency') + ylim(0,350)

ggplot(histDf, aes(x=nfriends,y=number)) +
  geom_point()  +
  xlab('Number of friends')  + 
  ylab('Frequency') 

ggplot(histDf, aes(x=logNfriends,y=logNumber)) +
  geom_point()  +
  xlab('log(Number of friends)')  + 
  ylab('log(Frequency)') 

ggplot(histDf, aes(x=nfriends,y=sqrtNumber)) +
  geom_point()  +
  xlab('Number of friends')  + 
  ylab('sqrt(Frequency)') 

```

Compute sampling distributions for mean for original and log-transformed data with n=30
```{r}
nRuns=1000
sampSize=30
sampleMeansOrig=replicate(5000,mean(sample(nfriends$Nfriends,sampSize)))
sampleMeansLog=replicate(5000,mean(sample(nfriends$logNfriends,sampSize)))
sampleMeansDf=data.frame(sampleMeansOrig,sampleMeansLog)
ggplot(sampleMeansDf,aes(sampleMeansOrig)) + geom_histogram(bins=100)
ggplot(sampleMeansDf,aes(sampleMeansLog)) + geom_histogram(bins=100)

```

A real world example: temporal discounting data from UH2.  

```{r}
discountData=read.csv('bickelhyplarge.csv',header=FALSE)
names(discountData)=c('subnum','discountRate')
discountData = discountData %>% mutate(logDiscountRate=log10(discountRate),
                                       sqrtDiscountRate=sqrt(discountRate),
                                       recipDiscountRate=1/discountRate)
ggplot(discountData,aes(discountRate)) + geom_histogram(bins=100)
ggplot(discountData,aes(logDiscountRate)) + geom_histogram(bins=100)
ggplot(discountData,aes(sqrtDiscountRate)) + geom_histogram(bins=100)
ggplot(discountData,aes(recipDiscountRate)) + geom_histogram(bins=100)

```

Get sampling distributions

```{r}
nRuns=1000
sampSize=30
sampleMeansOrig=replicate(5000,mean(sample(discountData$discountRate,sampSize)))
sampleMeansLog=replicate(5000,mean(sample(discountData$logDiscountRate,sampSize)))
sampleMeansDf=data.frame(sampleMeansOrig,sampleMeansLog)
ggplot(sampleMeansDf,aes(sampleMeansOrig)) + geom_histogram(bins=100)
ggplot(sampleMeansDf,aes(sampleMeansLog)) + geom_histogram(bins=100)

```

