---
title: "Simulation R lab - live coding"
output: html_notebook
---

```{r}
library(tidyverse)
set.seed(12345)
```

## 1. Generating random numbers

1. Generate 1000 random samples from a normal distribution with a mean of 100 and a standard deviation of 10

```{r}
mySamples <- rnorm(1000,mean = 100, sd = 10)
mySampleDf <- tibble(samples = mySamples)
```

Show the histogram of the data.

```{r}
ggplot(mySampleDf,aes(samples)) +
  geom_histogram(bins=10)
```

Compute the mean and standard deviation of those data

```{r}
mySampleDf %>% 
  summarise(myMean = mean(samples),
            mySd = sd(samples))
```

Add 10 to the data and see how it affects the mean and standard deviation.

```{r}

```

Multiply the data by 2 and see how it affects the mean and standard deviation

```{r}


```

## 2. Sampling with and without replacement

Generate a set of 10 unique random integers between 0 and 20, using a uniform distribution.

```{r}

```

Take a sample of size 8 from the dataset, without replacement.

```{r}

```

Take a sample of size 30 from the dataset, without replacement.

```{r}

```

Take a sample of 30 from the dataset, with replacement.

```{r}

```

## 3. Sum of squared errors

Compute the mean of the following numbers:

```{r}
testData <- c(1,2,3,4)

```

Compute the sum of squared errors from each of the data points.

```{r}

```

Compute the sum of squared errors from the mean

```{r}

```


## 4. Standard deviation and cumulative distributions

Generate a random sample of size 10,000 with a mean of zero and a standard deviation of 1.

```{r}

```


Plot a histogram and a cumulative distribution for these data.

```{r}

```

Add the cumulative density for the normal distribution

```{r}

```

What proportion of data points fall within the mean +/- 1 standard deviation?

```{r}

```

What about 2 standard deviations?

```{r}

```


## 5. Sampling error and standard error of the mean

Create a function that takes a sample of a particular size from a standard normal distribution (mean = 0, standard deviation = 1) and returns the mean of the sample.

```{r}

```

Now run that function 1000 times (using sampleSize = 30) and take the standard deviation of the resulting means.

```{r}

```

Compare this value to the theoretical standard error of the mean:

$$
SEM = \frac{SD}{\sqrt{n}}
$$

```{r}

```

## 6. Central limit theorem

Take 1000 samples from a uniform distribution and plot their histogram.

```{r}

```

Adapt the function from the previous section to use a uniform distribution instead of a normal.

```{r}

```

Run the function 1000 times and plot the histogram of the resulting means.  Try changing the sample size and see how this affects the shape of the sampling distribution.

```{r}

```


## 7. Confidence intervals

Using the sample from section 1, compute the 95 percent confidence interval for the mean, using the normal distribution.

```{r}

```


## 8. Bootstrapping

Using the same data as previous problem, compute the 95 percent confidence interval for the mean using bootstrapping.  

```{r}

```


