---
title: "Simulation R lab - live coding"
output: html_notebook
---

```{r}
library(tidyverse)
set.seed(12345)
```

## 1. Generating random numbers

1. Generate 1000 random samples from a normal distribution with a mean of 100 and a standard deviation of 10

```{r}
mySamples <- rnorm(1000,mean = 100, sd = 10)
mySampleDf <- tibble(samples = mySamples)
```

Show the histogram of the data.

```{r}
ggplot(mySampleDf,aes(samples)) +
  geom_histogram(bins=10)
```

Compute the mean and standard deviation of those data

```{r}
mySampleDf %>% 
  summarise(myMean = mean(samples),
            mySd = sd(samples))
```

Add 10 to the data and see how it affects the mean and standard deviation.

```{r}
mySampleDf %>% 
  mutate(samples = samples + 10) %>%
  summarise(myMean = mean(samples),
            mySd = sd(samples))

```

Multiply the data by 2 and see how it affects the mean and standard deviation

```{r}
mySampleDf %>% 
  mutate(samples = samples*2) %>%
  summarise(myMean = mean(samples),
            mySd = sd(samples))


```

## 2. Sampling with and without replacement

Generate a set of 10 unique random integers from 1 to 20, using a uniform distribution.

```{r}
myIntegers <- sample.int(20,size = 10,
                         replace = FALSE)
myIntegerDf <- tibble(samples = myIntegers)
```

Take a sample of size 8 from the dataset, without replacement.

```{r}
myIntegerSampleNoReplace <- myIntegerDf %>%
  sample_n(8,replace = FALSE)
unique(myIntegerSampleNoReplace)

```

Take a sample of size 30 from the dataset, without replacement.

```{r}
myIntegerSampleNoReplace <- myIntegerDf %>%
  sample_n(30,replace = FALSE)

```

Take a sample of 30 from the dataset, with replacement.

```{r}
myIntegerSampleWithReplace <- myIntegerDf %>%
  sample_n(30,replace = TRUE)

```

## 3. Sum of squared errors

Compute the mean of the following numbers:

```{r}
testData <- c(1,2,3,4)

testDataMean <- mean(testData)

testDataMean

```

Compute the sum of squared errors from each of the data points.

```{r}
# THIS IS BAD FORM!

sse1 <- sum((testData[1] - testData)^2)

sse2 <- sum((testData[2] - testData)^2)

sse3 <- sum((testData[3] - testData)^2)

sse4 <- sum((testData[4] - testData)^2)

```

Compute the sum of squared errors from the mean

```{r}

sseMean <- sum((testDataMean - testData)^2)
```


## 4. Standard deviation and cumulative distributions

Generate a random sample of size 10,000 with a mean of zero and a standard deviation of 1.

```{r}
myNormalSamples <- rnorm(10000)
myNormalSamplesDf <- tibble(
  samples = myNormalSamples)
```


Plot a histogram for these data.

```{r}
ggplot(myNormalSamplesDf,aes(samples)) +
  geom_histogram(bins=50)
```

Plot a cumulative distribution for these data.

```{r}

ggplot(myNormalSamplesDf,aes(samples)) +
  stat_ecdf()
  
```

Add the cumulative density for the normal distribution

```{r}

```

What proportion of data points fall within the mean +/- 1 standard deviation?

```{r}
myMean <- mean(myNormalSamplesDf$samples)
mySd <- sd(myNormalSamplesDf$samples)

myNormalSamplesDf <- myNormalSamplesDf %>%
  mutate(lessthan1sd = samples < (myMean - mySd),
         greaterthan1sd = samples > (myMean + mySd),
         inside1sd = !lessthan1sd & !greaterthan1sd)

mean(myNormalSamplesDf$inside1sd)
```

What about 2 standard deviations?

```{r}
myNormalSamplesDf <- myNormalSamplesDf %>%
  mutate(lessthan2sd = samples < (myMean - 1.96*mySd),
         greaterthan2sd = samples > (myMean + 1.96*mySd),
         inside2sd = !lessthan2sd & !greaterthan2sd)

mean(myNormalSamplesDf$inside2sd)

```


## 5. Sampling error and standard error of the mean

Create a function that takes a sample of a particular size from a standard normal distribution (mean = 0, standard deviation = 1) and returns the mean of the sample.

```{r}

mySampleMean <- function(sampleSize){
  mySample <- rnorm(sampleSize)
  return(mean(mySample))
}

```

Now run that function 1000 times (using sampleSize = 30) and take the standard deviation of the resulting means.

```{r}

```

Compare this value to the theoretical standard error of the mean:

$$
SEM = \frac{SD}{\sqrt{n}}
$$

```{r}

```

## 6. Central limit theorem

Take 1000 samples from a uniform distribution and plot their histogram.

```{r}

```

Adapt the function from the previous section to use a uniform distribution instead of a normal.

```{r}

```

Run the function 1000 times and plot the histogram of the resulting means.  Try changing the sample size and see how this affects the shape of the sampling distribution.

```{r}

```


## 7. Confidence intervals

Using the sample from section 1, compute the 95 percent confidence interval for the mean, using the normal distribution.

```{r}

```


## 8. Bootstrapping

Using the same data as previous problem, compute the 95 percent confidence interval for the mean using bootstrapping.  

```{r}

```


