---
title: "Memory recall analysis"
output:
  html_document:
    df_print: paged
---

# Environment set up
```{r}
#Libraries
library(tidyverse)

#File paths
respData_path <- "https://raw.githubusercontent.com/psych10/psych10/master/exercises/memoryRecall/Memory%20recall%20test.csv"
```

# Read in data 
```{r}
respData <- 
  read_csv(respData_path) %>% 
  mutate(
    nRecalled = NULL,
    nHits = 0
  )
```

## Analyze
```{r}
# last item in this list is the DRM critical lure

listWords <- c(
  "fear", "temper", "hatred", "fury", "happy",
  "enrage", "emotion", "rage", "hate", "mean",
  "ire", "mad", "wrath", "fight", "anger"
)

itemData <- c()

for (i in 1:dim(respData)[1]) {
  split_resp <- strsplit(tolower(respData[i, 2]), " ")[[1]]
  respData$nRecalled[i] <- length(split_resp)

  for (j in 1:length(listWords)) {
    itemData <- rbind(itemData, c(j, as.integer(listWords[j] %in% split_resp)))
  }
}

itemDataDf <- as.data.frame(itemData)
names(itemDataDf) <- c("position", "hit")
```

# Summarize the data in order to plot it
```{r}
#create a dataframe that summarizes the hit rates for each word
sm <- 
  itemDataDf %>% 
  group_by(position) %>% 
  summarize(
    n = n(),
    mean_hit = mean(hit),
    sd_hit = sd(hit),
    se_hit = sd_hit / sqrt(n)
  )
  
#add a new variable specify whether each word was or was not the lure 
itemDataDf <- 
  itemDataDf %>%
  mutate(
    lure = as.integer(
      position == length(listWords)
    )
  )

#calculate the mean hit rate for the lure
lureHitRate <- 
  itemDataDf %>% 
  filter(lure == 1) %>% 
  summarize(
    lureHitRate = mean(hit)
  ) %>% 
  pull() #pull() saves "lureHitRate" to a single value rather than keeping it in a dataframe
```

#Plot serial position curve

## Method 1
Basic plot without word label information. 
```{r}
sm %>% 
  filter(position < length(listWords)) %>% 
  ggplot(aes(x = position, y = mean_hit)) + 
  geom_line() +
  geom_hline(yintercept = lureHitRate, color = "blue") +
  scale_x_continuous(breaks = 1:14) +
  ylim(0, 1) +
  annotate(
    "text", 
    label = "hit rate for lure ('anger')", 
    x = 7, 
    y = lureHitRate * 0.85, 
    colour = "blue"
  ) +
  labs(
    x = "Serial position of item in list",
    y = "Proportion correct recall"
  )
```

## Method 2
Use `scale_x_continuous` to replace the numbers with the words on the x-axis.
```{r}
sm %>% 
  filter(position < length(listWords)) %>% 
  ggplot(aes(x = position, y = mean_hit)) + 
  geom_line() +
  geom_hline(yintercept = lureHitRate, color = "blue") +
  scale_x_continuous(
    breaks = 1:(length(listWords) - 1), #create a sequence of breaks of length 14 (because not plotting the lure word "anger" on the x-axis)
    labels = listWords[1:(length(listWords) - 1)]
  ) + 
  ylim(0, 1) +
  annotate(
    "text", 
    label = "hit rate for lure ('anger')", 
    x = 7, 
    y = lureHitRate * 0.85, 
    colour = "blue"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, vjust = 0.5)
  ) +
  labs(
    x = "Serial position of item in list",
    y = "Proportion correct recall"
  )
```

## Method 3
Use `geom_text` to add the words as labels to the points. 
```{r}
sm %>% 
  filter(position < length(listWords)) %>% 
  ggplot(aes(x = position, y = mean_hit)) + 
  geom_line() +
  geom_hline(yintercept = lureHitRate, color = "blue") +
  geom_label( #new
    label = listWords[1:(length(listWords) - 1)]
  ) +
  scale_x_continuous(
    breaks = 1:14
  ) + 
  ylim(0, 1) +
  annotate(
    "text", 
    label = "hit rate for lure ('anger')", 
    x = 9, 
    y = lureHitRate * 0.85, 
    colour = "blue"
  ) +
  labs(
    x = "Serial position of item in list",
    y = "Proportion correct recall"
  )
```

## Method 4
The functions in the additional package `ggrepel` sometimes do a bit better of a job positioning the labels for you.  There are also arguments for both `geom_text` and `geom_label_repel` than can nudge the position of the labels. 

We've also added `geom_point` here to make things a bit clearer.
```{r}
#install.packages("ggrepel")
library(ggrepel)

sm %>% 
  filter(position < length(listWords)) %>% 
  ggplot(aes(x = position, y = mean_hit)) + 
  geom_point() +
  geom_line() +
  geom_hline(yintercept = lureHitRate, color = "blue") +
  geom_label_repel( #new
    label = listWords[1:(length(listWords) - 1)]
  ) +
  scale_x_continuous(
    breaks = 1:14
  ) + 
  ylim(0, 1) +
  annotate(
    "text", 
    label = "hit rate for lure ('anger')", 
    x = 9, 
    y = lureHitRate * 0.85, 
    colour = "blue"
  ) +
  labs(
    x = "Serial position of item in list",
    y = "Proportion correct recall"
  )
```
