---
title: "Psych 10 Project: Milestone 1"
output: html_notebook:
   toc: true
   toc_float: true
---

NOTE: Submit this file as a .Rmd file on Canvas. 

This is Part 1 of your project. Ultimately, you will integrate Part 1 with a final project write up (10% of your grade) and presentation (5% of your grade). The goals of Part 1 of your project are to:

* Identify your question and explain why you are asking it
* Identify, briefly describe, and read in your dataset to R
* Identify and descdribe the variables you will use to test your hypothesis.
* Identify challenges of this dataset and what data wrangling/cleaning you need to do. 
* Identify limitations of your project as a whole. 

# Environment set up 
```{r}
#Libaries
#Load the packages you need here

#Files
#save the URL path that we provided to your data file 
# e.g., data_file <- "path"
```

# Project question

__q1__ What question will you answer in your project? 




__q2__ Motivation: Why is this question interesting and/or important to ask?




__q7__ Phrase your question in terms of a testable hypothesis. 



# The dataset

__q1__ What publically available dataset will you use for your project? Provide a brief description of this dataset (e.g., who collected it and when, what data does it contain, why was it collected). 



__q2__ What file format is this dataset (e.g., .csv, .txt, .rds, .xlxs, data within R package)?



__q3__ Using one of the `readr`, `haven`, or `read_excel` functions to read in the data file and save it is a data frame in your R Studio Environment. 

```{r}

```

__q4__  Use the glimpse() function take a look at your dataset. 
```{r}

```

__q5__ How many observations are in your dataset and how many variables? 


__q6__ Which variables will use to to test your hypothesis? Describe what they are. Include what types of variables they are (e.g., character, integer, etc.) and what their scales of measurement are. 


# Data science challenges

_q1_ Is the raw dataset "tidy" (https://r4ds.had.co.nz/tidy-data.html)? Why or why not?


_q2_ What will you do to clean and wrangle your dataset to make it appropriate for analysis? For example, will you remove certain columns or rows? Do you need to rename or recode variables? 


# Project limitations

__q1__ What are some of the limitations of your data that may make it challenging to draw particular conclusions or interpret your findings (e.g., considering who is in the sample and the design of the study)?


# Team member contributions

__q1__ Please explain how each of your team members (by name) contributed to Part 1 of your project. 








